The complex network of trophic interactions in a subAntarctic Marine Protected

### 2023-02-06 MS_ToSubmit/MS_Burdwood_foodweb.Rmd
Finish 1st draft

Food web connectance is a feature that resumes the complexity of the network, but more importantly it is an emergent property of pairwise species interactions [@Poisot2014]. It contains information regarding how interactions within an ecological network are distributed, and predicts reasonably well key dynamical properties of ecological networks [@Dunne2002a].
Complex marine food webs (i.e. with more than 25 trophic species) show connectance values ranging from 0.01 - 0.27 [@Marina2018a]. In particular, food webs from high-latitude regions tend to exhibit a connectance closer to the minimum (between 0.01 and 0.05) [@Santana2013; @Kortsch2015; @Rodriguez2022]. Whether food webs display a low or a high connectance helps to better comprehend ecosystem's synthetic properties like robustness. In this sense, empirical analyses support the notion that highly-connected ecological networks are robust against external perturbations such as the introduction of new (e.g., invasive) species [@Smith-Ramesh2017] as well as species removal (e.g., local extinction) [@Dunne2002; @Montoya2003]. The connectance of the food web of the MPA Namuncurá - Banco Burdwood (0.01) is one of the lowest reported so far for these regions; five times lower than that of Beagle Channel [@Rodriguez2022].

The degree distribution or the distribution of number of interactions per species is the core for the structure of species interactions, which influences the opportunities for multiple species to persist in the long term and, therefore, their coexistence [@Godoy2018]. The food web for the MPA N-BB presents an asymmetric degree (interaction) distribution. This is a pattern that we have identified at different levels of analysis: food web, predator, prey, and functional group. Such asymmetry is a well-known feature in empirical complex food webs in particular [@Dunne2002a; @Montoya2003; @Stouffer2005], and has received great attention in complex networks in general [@Albert2002; @Newman2003].
The degree distribution affects the resilience of complex food webs against random failures and intentional attacks: right-skewed distributions, like the one described in this study, are more vulnerable to the removal of the most-connected species or hubs with the potential of producing secondary extinctions and a catastrophic fragmentation of the network [@Albert2000; @Dunne2002; @Eklof2006].

It is suggested that the small-world pattern is not frequent in complex marine food webs, mainly due to a low clustering coefficient compared to random networks [@Dunne2002b; @Marina2018a]. However, the food web of the MPA N-BB does display a small-world pattern, characterizing a network with short path length and high clustering coefficient. Consequences of the small-world pattern in food webs are of great importance in recognizing evolutionary paths and the vulnerability to perturbations [@Montoya2002]. On the one hand, a short path length implies a rapid spread of an impact (e.g., contaminant, population fluctuation, local extinction) throughout the network and, at the same time, a more potentially adaptive dynamics in face of external perturbations [@Montoya2002; @Williams2002]. On the other hand, a high clustering coefficient indicates the formation of subnetworks composed only by the neighbors of particular species. This translates into a greater resistance of the network due to the confinement of perturbations mainly within subnetworks and not between them [@Kortsch2019; @Heer2020]. Overall, a small-world topology provides ecological networks with a greater resilience and resistance [@Bornatowski2017; @Dormann2017].

@@

Overall, the food web of the MPA Namuncurá - Banco Burdwood presents a combination of network properties that make it unique in terms of network resolution, complexity, and structural pattern. This suggests that the ecosystem might be fragile to external perturbations targeting high connected species. However, structural properties might provide resilience and resistance with the final outcome of a rearrange structure maintaining their functions.

@@

The degree distribution, a network-level property as defined in this study, also allows to identify important species, such as potential keystone species [@Sole2001; @Dunne2002], generalists/specialist species, and dominant food sources [@Kondoh2010].

We have identified that the majority of the consumers in the food web of the MPA N-BB have a narrow diet or are specialists, while few present a broad diet. *Patagonotothen guntheri*, *P. ramsayi*, *Dissostichus eleginoides*, *Bathyraja brachyurops*, and *B. griseocauda* are the most evident generalist species, with more than 25 potential prey.
Since these species present mid-trophic positions in the food web (3-4), acting as predator and prey, they might be important links between lower and higher trophic levels. This is in agreement with the sole analysis, using stable isotopes, that exists so far for the trophic structure of the MPA N-BB [@Riccialdelli2020], and resembles other high-latitude marine systems of the Southwest Atlantic and Arctic regions [@Arkhipkin2013; @Kortsch2019]. The importance of these generalist species also arises since they feed on the benthic and pelagic habitats [@Troccoli2020; @CovattiAle2022; @Fischer2022], linking these realms and contributing to the vertical carbon flow.

On the other hand, few prey are consumed by many predators in the food web of the MPA N-BB. This suggests that there are dominant food sources from which most consumers depend on, and from where the ecosystem energy is being transferred to the upper trophic levels. The most demanded source we identified in this study (i.e. detritus) supports the abundant benthic community of filter-feeders [@Schejter2016]. Furthermore, we found that the second and third-most consumed prey were diatoms and species of *Euphausia*, which are essential sources for the diverse zooplankton community on one hand [@Spinelli2020], mid-TL consumers like the Fuegian sprat *Sprattus fuegensis* [@Padovani2021] and *Patagonotothen ramsayi* [@Fischer2022], and top predators on the other hand [@RayaRey2022; @Riccialdelli2022].

@@

Closeness and betweenness are defined as "meso-scale" properties because they consider direct and indirect interactions, therefore describing the focal species’ ability to influence the rest of the species of the food web [@Lai2012].
Closeness quantifies how many steps away species $i$ is from all others in the food web, and is proportional to how rapidly the indirect effects of the focal species can spread to other species in the network [@Scotti2010]. In the food web of the MPA N-BB, low-TL consumers arise as important in this regard: species of the zooplankton community, *Calanus* and *Euphausia*, and Brachiopoda. Any perturbation affecting these species, such as the already confirmed contaminants mercury [@Fioramonti2022] and microplastics [@Cossi2021; @DiMauro2022] for the area, should be of concern since it might reach many other species in the food web.
Otherwise, betweenness measures the number of shortest paths between species, providing information on the importance of species as “bridges” for energy transfer: a species with high betweenness takes part in more food chains and therefore affects more energy flows [@Scotti2010]. We have identified the longtail southern cod *Patagonotothen ramsayi* as the most important species in this sense. Moreover, according to our analysis, other species like the Patagonian toothfish *Dissostichus eleginoides*, the Patagonian cod *Salilota australis*, the yellowfin notothen *Patagonotothen guntheri*, and the Patagonian longfin squid *Doryteuthis gahi* should be consider as relevant energy transfers for the ecosystem. All these species have a mid-trophic position in the food web, supporting the 'wasp-waist' control hypothesis for the MPA N-BB [@Riccialdelli2020].

Ecosystems with a pronounced 'wasp-waist' structure are suggested to present a high trophic redundancy, since a large proportion of species would show similar trophic habits [@Cury2000]. The significant negative relationship we found between trophic similarity and trophic level enhances the hypothesis of functional redundancy at low and mid-TL species compared to higher TL species for the MPA N-BB food web [@Riccialdelli2020]. At the same time, our results highlight the uniqueness in terms of trophic role for high-TL predators. Here, not only the expected pelagic animals as marine mammals and sea birds arise as relevant, but also benthic vertebrate (chondrichthyans *Bathyraja macloviana* and *Squalus acanthias*) and invertebrate species (echinoderms *Diplopteraster clarki* and *Pteraster sp*) are noteworthy. The role that such species play in the MPA N-BB ecosystem is unique and perturbations on them might result in unprecedented changes at the trophic structure and functioning.

Species' role can be also assessed in a module-based context, where modules are defined as a group of species that interact more frequently among themselves than with species that are not members of the module [@Guimera2005]. Among the varying numbers of topological roles in which species can be divided, two are remarkable: 'module connector' and 'network connector'. Here again the longtail southern cod *Patagonotothen ramsayi* is emphasized as an influential species for the connectivity of the food web, which agrees with the suggested 'wasp-waist' control for the MPA N-BB ecosystem [@Riccialdelli2020]. But most importantly, our analysis broadens this issue by pointing out that there are actually other species from a wide range of trophic positions (1 to 3.86), representing 20 different functional groups, that should be considered. Thus, we propose that it is the diversity of species (benthic and pelagic) what maintains the connectivity of the food web, therefore contributing to the trophic structure and ecosystem's stability.

## Caveats and future perspectives

Caveats:

Seasonal variation in predators' diet due to food source availability.
@Brose2019

Spatial variation of study area: Banco Burdwood west-east gradient.
In the oceanic BB area, complex oceanographic conditions generate an internal spatial heterogeneity, mainly along its longitudinal axis [@Matano2019], and consequently plankton communities respond with different spatial distributions [@Spinelli2020].
@Kortsch2019

The lack of density or biomass data might have neglected well-known important species for the structure of the food web: *Sprattus fuegensis*.

Perspectives:

Incorporate habitat heterogeneity, since benthic community in the BB and deep flanks is different [@Schejter2021]. This might lead to distinct food web properties, hence differences in ecosystem functioning and response to anthropogenic and environmental perturbations.
@Cordone2020

Estimation of interaction strength considering species and interaction traits (i.e. body size, body mass, dimensionality) to gain insights into food web stability and response to anthropogenic and environmental perturbations.
@Nilsson2016; @Cirtwill2018; @Kortsch2021

Simulate perturbations on food web and target species: mercury transfer, microplastic pollution, fisheries.
@Cossi2021; @DiMauro2022; @Fioramonti2022

@@

@@

This study was funded by Consejo Nacional de Investigaciones Científicas y Técnicas (CONICET) and Agencia Nacional de Promoción Científica y Tecnológica (PICT 2020-SERIEA-01617), Argentina. This work is contribution no. XX of the MPA Namuncurá (Law 26875).

>>>>

Food web connectance is a feature that resumes the complexity of the network, but more importantly it is an emergent property of pairwise species interactions [@Poisot2014]. It contains information regarding how interactions within an ecological network are distributed, and predicts reasonably well key dynamical properties of ecological networks [@Dunne2002a].
Complex marine food webs (i.e. with more than 25 trophic species) show connectance values ranging from 0.01 - 0.27 [@Marina2018a]. In particular, food webs from high-latitude regions tend to exhibit a connectance closer to the minimum (between 0.01 and 0.05) [@Santana2013; @Kortsch2015; @Rodriguez2022]. Whether food webs display a low or a high connectance helps to better comprehend ecosystem's synthetic properties like robustness. In this sense, empirical analyses support the notion that highly-connected ecological networks are robust against external perturbations such as the introduction of new (e.g., invasive) species [@Smith-Ramesh2017] as well as species removal (e.g., local extinction) [@Dunne2002; @Montoya2003]. The connectance of the food web of the MPA Namuncurá - Banco Burdwood (0.01) is one of the lowest reported so far for these regions; five times lower than that of Beagle Channel, an adjacent area [@Rodriguez2022].

The degree distribution or the distribution of number of interactions per species is the core for the structure of species interactions, which influences the opportunities for multiple species to persist in the long term and, therefore, their coexistence [@Godoy2018]. The food web for the MPA N-BB presents an asymmetric degree distribution. This is a pattern that we have identified at different levels of analysis: food web, predator, prey, and functional group. Such asymmetry is a well-known feature in empirical complex food webs in particular [@Dunne2002a; @Montoya2003; @Stouffer2005], and has received great attention in complex networks in general [@Albert2002; @Newman2003].
The degree distribution affects the resilience of complex food webs against random failures and intentional attacks: right-skewed distributions, like the one described in this study, are more vulnerable to the removal of the most-connected species or hubs with the potential of producing secondary extinctions and a catastrophic fragmentation of the network [@Albert2000; @Dunne2002; @Eklof2006].

It is suggested that the small-world pattern is not frequent in complex marine food webs, mainly due to a low clustering coefficient compared to random networks [@Dunne2002b; @Marina2018a]. However, the food web of the MPA N-BB does display a small-world pattern, characterizing a network with short path length and high clustering coefficient. Consequences of the small-world pattern in food webs are of great importance in recognizing evolutionary paths and the vulnerability to perturbations [@Montoya2002]. On the one hand, a short path length implies a rapid spread of an impact (e.g., contaminant, population fluctuation, local extinction) throughout the network and, at the same time, a more potentially adaptive dynamics in face of external perturbations [@Montoya2002; @Williams2002]. On the other hand, a high clustering coefficient indicates the formation of subnetworks composed only by the neighbours of particular species. This translates into a greater resistance of the network due to the confinement of perturbations mainly within subnetworks and not between them [@Kortsch2019; @Heer2020]. Overall, a small-world topology provides ecological networks with a greater resilience and resistance [@Bornatowski2017; @Dormann2017].

@@

To resume, the food web of the MPA Namuncurá - Banco Burdwood presents a combination of network properties that make it unique in terms of network resolution, complexity, and structural pattern. All this suggest that the ecosystem might be fragile to external perturbations targeting high connected species. However, structural properties might provide resilience and resistance with the final outcome of a rearrange structure maintaining their functions.

@@

The degree distribution allows to identify important species, such as potential keystone species [@Sole2001; @Dunne2002], generalist/specialist species, and dominant food sources [@Kondoh2010].

We have identified that the majority of the consumers in the food web of the MPA N-BB has a narrow diet or are specialists, while few present a broad diet. The most evident generalist species are *Patagonotothen guntheri*, *P. ramsayi*, *Dissostichus eleginoides*, *Bathyraja brachyurops*, and *B. griseocauda*, with more than 25 potential prey.
Since these species present mid-trophic positions in the food web (3-4), acting as predator and prey, they might be important links between lower and higher trophic levels. This is in agreement with the sole analysis, using stable isotopes, that exists so far for the trophic structure of the MPA N-BB [@Riccialdelli2020], and resembles other high-latitude marine systems of the Southwest Atlantic and Arctic regions [@Arkhipkin2013; @Kortsch2019]. The importance of these particular generalist species also arises since they feed on the benthic and pelagic habitats [@Troccoli2020; @CovattiAle2022; @Fischer2022], linking these realms and contributing to the vertical carbon flow.

On the other hand, few prey are consumed by many predators in the food web of the MPA N-BB. This suggests that there are dominant food sources from which most consumers depend on, and from where the ecosystem energy is being transferred to the upper trophic levels. The most demanded source we identified in this study (i.e. detritus) supports the abundant benthic community of filter-feeders [@Schejter2016]. Furthermore, we found that the second and third-most consumed prey were diatoms and species of *Euphausia*, respectively, which are essential sources for the diverse zooplankton community [@Spinelli2020], mid-TL consumers like the Fuegian sprat *Sprattus fuegensis* [@Padovani2021] and *Patagonotothen ramsayi* [@Fischer2022], and top predators [@RayaRey2022; @Riccialdelli2022].

@@

Closeness and betweenness are defined as "meso-scale" properties because they consider direct and indirect interactions, therefore describing the focal species’ ability to influence the rest of the species of the food web [@Lai2012].
Closeness quantifies how many steps away species $i$ is from all others in the food web, and is proportional to how rapidly the indirect effects of the focal species can spread to other species in the network [@Scotti2010]. In the food web of the MPA N-BB, low-TL consumers arise as important in this regard: species of the zooplankton community, *Calanus* and *Euphausia*, and Brachiopoda. Any perturbation affecting these species, such as the recently confirmed contaminants mercury [@Fioramonti2022] and microplastics [@Cossi2021; @DiMauro2022] for the area, should be of concern since it might reach many other species in the food web.
Otherwise, betweenness measures the number of shortest paths between species, providing information on the importance of species as “bridges” for energy transfer: a species with high betweenness takes part in more food chains and therefore affects more energy flows [@Scotti2010]. We have identified the longtail southern cod *Patagonotothen ramsayi* as the most important species in this sense. Moreover, in light of our analysis, other species like the Patagonian toothfish *Dissostichus eleginoides*, the Patagonian cod *Salilota australis*, the yellowfin notothen *Patagonotothen guntheri*, and the Patagonian longfin squid *Doryteuthis gahi* should be consider as relevant energy transfers for the ecosystem. All these species have a mid-trophic position in the food web, supporting the 'wasp-waist' control hypothesis for the MPA N-BB [@Riccialdelli2020].

Ecosystems with a pronounced 'wasp-waist' structure are suggested to present a high trophic redundancy, since a large proportion of species would show similar trophic habits [@Cury2000]. The significant negative relationship we found between trophic similarity and trophic level enhances the hypothesis of functional redundancy at low and mid-TL species compared to higher TL species for the MPA N-BB food web [@Riccialdelli2020]. At the same time, our results highlight the uniqueness in terms of trophic role for high-TL predators. Here, not only the expected pelagic animals such as marine mammals and sea birds arise as relevant, but also benthic vertebrate (chondrichthyans *Bathyraja macloviana* and *Squalus acanthias*) and invertebrate species (echinoderms *Diplopteraster clarki* and *Pteraster sp*) are noteworthy. The role that such species play in the MPA N-BB ecosystem is unique and perturbations on them might result in unprecedented changes at the trophic structure and functioning.

Species' role can be also assessed in a module-based context, where modules are defined as a group of species that interact more frequently among themselves than with species that are not members of the module [@Guimera2005]. Among the varying numbers of topological roles in which species can be divided, two are remarkable: 'module connector' and 'network connector'. Here again the longtail southern cod *Patagonotothen ramsayi* is emphasized as an influential species for the connectivity of the food web, which agrees with the suggested 'wasp-waist' control for the MPA N-BB ecosystem [@Riccialdelli2020]. But most importantly, our results broadens this issue by pointing out that there are other species, belonging to a wide range of trophic positions (1 to 3.86) and representing 20 different functional groups, that should be considered. Thus, we propose that it is the diversity of species (benthic and pelagic) what maintains the connectivity of the food web, therefore contributing to the trophic structure and ecosystem's stability.

<!-- ## Caveats and future perspectives -->

<!-- Caveats: -->

<!-- Seasonal variation in predators' diet due to food source availability. -->
<!-- @Brose2019 -->

<!-- Spatial variation of study area: Banco Burdwood west-east gradient. -->
<!-- In the oceanic BB area, complex oceanographic conditions generate an internal spatial heterogeneity, mainly along its longitudinal axis [@Matano2019], and consequently plankton communities respond with different spatial distributions [@Spinelli2020]. -->
<!-- @Kortsch2019 -->

<!-- Perspectives: -->

<!-- Incorporate habitat heterogeneity, since benthic community in the BB and deep flanks is different [@Schejter2021]. This might lead to distinct food web properties, hence differences in ecosystem functioning and response to anthropogenic and environmental perturbations. -->
<!-- @Cordone2020 -->

<!-- Simulate perturbations on food web and target species: mercury transfer, microplastic pollution, fisheries. -->
<!-- @Cossi2021; @DiMauro2022; @Fioramonti2022 -->

<!-- Estimation of interaction strength considering species and interaction traits (i.e. body size, body mass, dimensionality) to gain insights into food web stability and response to anthropogenic and environmental perturbations. -->
<!-- @Nilsson2016; @Cirtwill2018; @Kortsch2021 -->

@@

We compiled information on the species and trophic diversity of the Marine Protected Area Namuncurá - Banco Burdwood, generating an unprecedented well-resolved network of trophic interactions for a subAntarctic ecosystem. Particular properties at the network level might reflect ecosystem's vulnerability to perturbations of highly-connected species, with a rearrange structure maintaining their functions due to its potential resilience and resistance.

We identified several species as important in terms of different aspects of trophic structure and functioning, and response to perturbations (i.e. environmental/anthropogenic changes). On one hand, we suggest that generalist species, mostly fishes, play a crucial role in the ecosystem's bentho-pelagic coupling process. At the same time, we propose that other species rather than the longtail southern cod *Patagonotothen ramsayi* and the Fueguian sprat *Sprattus fuegensis* should be consider as relevant energy transfers for the ecosystem. Finally, we argue that it is the diversity of species, representing the benthic and pelagic habitats, what maintains the connectivity of the food web against perturbations, therefore contributing to the structure and stability of the ecosystem.

@@

We are indebted to all those experts of the working group 'Banco Burdwood / Área Protegida Namuncurá' who humbly provided their knowledge to enhance the quality of the present research.
This study was funded by Consejo Nacional de Investigaciones Científicas y Técnicas (CONICET) and Agencia Nacional de Promoción Científica y Tecnológica (PICT 2020-SERIEA-01617), Argentina. This work is contribution no. XX of the MPA Namuncurá (Law 26875).
